---
title: "Sample size analysis with non informative prior"
author: "BZ"
date: "March 17, 2017"
header-includes: 
   - \usepackage{color}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message=FALSE, warning=FALSE)
```

## Sample size calculation using non informative prior

The choice of non-informaitve prior for binomial distribution resulting in posterior distribution see http://www.stats.org.uk/priors/noninformative/YangBerger1998.pdf

  + if we choose Beta(1,1), that is, the uniform distribution, then the posterior mean ADA+ rate would be (x + 1)/(n + 2) where x and n are the number of ADA+ incidence and the total number of subjects at cohort 1, respectively; 
  +  if we choose Jefferys' prior, then the posterior mean ADA+ rate would be (x + 0.5)/(n + 1).

In the following section, we'll base sample size re-estimation on these two non-informative priors. Note that the initial sample size 102 (with 46 in the first cohort) is calculated based on the assumption that ADA+ rate is 1%. 

The following is variable specifications for tables below:

  + n_ADA: number of ADA observed in cohort 1
  + n_cohort1: number of subjects enrolled in cohort 1
  + obs_ADA_rate: n_ADA/n_cohort1
  + posterior_ADA_rate: obtained using Bayes prior and data from cohort 1
  + N_required_per_arm: sample size re-estimated using posterior_ADA_rate. This number has already accouted for 10% drop out. 
  + N_increased_per_arm: number of subjects needed per arm (N_required_per_arm - 51, and 51 is sample size per arm originally).
  
  \textcolor{red}{Please note that these sample sizes may change due to the type 1 error analysis, addressing comment 3 from FDA.}

```{r}
library(gsDesign)
library(knitr)
n_prior <- function(alpha, beta, x = 0:7, n = 46, total = 102){

  p_posterior <- (alpha + x)/(alpha + beta + n)
  
  nsize <- nBinomial(p1 = p_posterior, p2= p_posterior, alpha  = 0.05, beta = 0.2, delta0 = -0.1, 
                     ratio = 1, sided = 1, outtype = 2)
  final_n <- ceiling(as.vector(ceiling(nsize)[, 1])*1.1)
  # additional
  addtional <- ceiling(final_n)-total/2
  addtional[addtional<0] <- 0
  result <- data.frame(n_ADA = x, n_cohort1 = n, obs_ADA_rate = round(x/n, 4), 
                       posterior_ADA_rate = round(p_posterior, 4), 
                       N_required_per_arm = final_n, N_increase_per_arm = addtional)
  return(result)
}

```
\newpage

# when initial ADA rate is 1%

In this part, the initial sample size 102 (with 46 in the first cohort) is calculated based on the assumption that ADA+ rate is 1%. 

## Uniform prior
```{r}
alpha <- beta <- 1
kable(n_prior(alpha=1, beta=1))
```



## Jeffery's prior
```{r}
alpha <- 1/2; beta <- 1/2
kable(n_prior(alpha=1/2, beta=1/2))
```

## Reproducing Table 1 using our package

```{r}
kable(n_prior(alpha =1, beta = 100))
```

\newpage
# when initial ADA rate is 3%

In this part, the initial sample size 132 (with 60 in the first cohort) is calculated based on the assumption that ADA+ rate is 3%. 



## Uniform prior
```{r}
alpha <- beta <- 1
kable(n_prior(alpha=1, beta=1, n = 60, total = 132))
```


## Jeffery's prior
```{r}
alpha <- 1/2; beta <- 1/2
kable(n_prior(alpha=1/2, beta=1/2, n = 60, total = 132))
```


\newpage
# Posterior distribution based on Jeffreys prior and 1% initial ADA rate

```{r}
library(dplyr)

dens <- function(x, n, p){
  prob <- dbeta(p, shape1 = x + 1/2, shape2 = n - x + 1/2)
  data <- data.frame(p = p, density = prob)
  return(data)
}
 p <- seq(0.01, 0.99, by = 0.001)    # 
  prior <- 1/pi*p^(-0.5)*(1-p)^(-0.5)
  dat <-  data.frame(p = p, density = prior, case = " Prior") 
  n <- 46
  p1 <- dens(x = 1, n = 46, p = p) %>% mutate(case = "Posterior Est from the Interim Rate = 1/46")
  p2 <- dens(x = 3, n = 46, p = p) %>% mutate(case = "Posterior Est from the Interim Rate = 3/46")
  p3 <- dens(x = 5, n = 46, p = p) %>% mutate(case = "Posterior Est from the Interim Rate = 5/46")
  

  den <- bind_rows(dat, p1, p2, p3)
  
library(ggplot2)
  
  ggplot(data = den, aes(x = p, y = density)) + geom_line(aes(color = case)) + 
      facet_wrap(~case, scales = "free") + 
      theme(legend.position = "none")
  
  

```
The top left is the Jeffreys prior. The top right shows the posterior distribution Beta(1.5, 45.5); 
The lower left Beta(3.5, 43.5) and the lower right Beta(5.5, 41.5).











