---
title: "Three dots"
author: "Bin Zhuo"
date: "February 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this document, I'm going to explore the three-dot `...` that is frequently used in R. An example would be
`print(x, ...)`.


## The three dots in **R**
In a technical language, this `...` is called an **ellipsis**. It means that the function is designed to take any number of named or unnamed arguments. This ellipsis is not exclusively seen in **R**, and other programming languages have a similar concept. 

The first question is: how would you make use of ellipsis when writing functions? An example is provided below
```{r}
helloworld <- function(...){
  arguments <- list(...)
  print((arguments))
}

helloworld("I'm stupid!")
helloworld("I'm stupid!", "I'm very stupid!")
helloworld("I'm stupid!", "I'm very stupid!", "I'm very very stupid!")
```
Now probably you've got a sense of what `helloworld()` function does: it literaly converts whatever you put in the function as a list, and then prints the list. 


## So, when should I use the ellipsis?
There are essentially two situations where you can use the three dots:
  
  + When it makes sense to call the function with **a variable number of arguments**, as we've seen from the `helloworld()` function. 
  + When, within your function, you call other functions, and these functions can have a variable number of arguments. A simple example is provided
  
  
```{r}
fun1 <- function(x, ...){
  y <- mean(x, ...)
  return(y)
}
x <- c(1, 2, NA, 4)
fun1(x, na.rm = TRUE)
fun1(x, na.rm = FALSE)
```
In the above example, the called function `mean()` is generic, therefore `...` allows us to whether `NA` should be excluded or not from calculation. In other cases, the called function could be passed into the function as an argument. Here's an example:

```{r}
fun2 <- function(a, b){
  
  x <- (a + b)/2
  return(x)
}

fun3 <- function(c, ...){
  y1 <- fun2(...)
  y2 <- (c + y1)/2
  return(y2)
}

fun3(4, 2, 1)
```



